Dynamical modeling of tidal streams

### 2014-01-06 tex/ms.tex
streams: tweak to appendix

Recently, Fox \& Binney (2014, in preparation) have suggested a new
method for calculating $\vecj(\vecx,\vecv)$ inspired by torus
modeling. This method requires only a single orbit integration and
allows the actions to be calculated accurately and as precisely as
desired (that is, a convergence criterion can be applied and
convergence can be attained). I describe this method briefly here and
then show how to extend it to also calculate the frequencies and
angles: $(\vecx,\vecv) \rightarrow (\veco,\veca)$.

@@

The Fox \& Binney method consists of averaging \eqnname~(\ref{eq:jjt})
over the auxiliary angle along the path of the orbit, such that the
mapping defined by the $S_{\vecn}$ is onto orbital tori of the target
Hamiltonian, the oscillating cosine behavior cancels, and the target
action is obtained as the auxiliary-angle-averaged auxiliary
action. That is,
\begin{equation}\label{eq:jjtint}
  \int \dd \theta_i^A J_i^A = \int \dd \theta_i^A J_i + 2 \sum_{\vecn > 0} \vecn\,S_{\vecn}(J_i)\int \dd \theta_i^A\cos(\vecn\cdot\veca^A)\,,
\end{equation}
where the integral is over auxiliary angles calculated on the path of
the orbit in the target potential. The factor $\vecn\,S_{\vecn}(J_i)$
comes out of the integral as the target actions are conserved along
the orbit. In this Equation, we integrate over the angle coordinate
$\theta_i^A$ associated with the action $J_i$ that we want to compute
(that is, $\theta_R$ for $J_R$ and so on). This is because of the
factor $\vecn$: the condition $\vecn > 0$ means that when calculating
$J_R$ there are terms with $\vecn = (n_R,0,0)$, which would not cancel
according to the argument below when integrating over $\theta_Z$. For
non-resonant orbits and after a long-enough orbit integration, the
angles can be considered to be independent (that is, the orbit ``fills
its torus''), such that we can write (for example, for $\theta_R$)
\begin{equation}
\begin{split}
  \int \dd \theta_R^A\cos(\vecn\cdot\veca^A) & = \mathrm{Re}\left[\int \dd \theta_R^A\,\exp(-i\vecn\cdot\veca^A)\right]\,, \\
  & \approx \mathrm{Re}\left[\exp(-i\,n_\phi\,\theta^A_\phi-i\,n_Z\,\theta^A_Z)\int \dd \theta_R^A\,\exp(-i\,n_R\,\theta^A_R)\right]\,.
\end{split}
\end{equation}
As long as the auxiliary angle $\theta_R$ goes through the full range
$0$ to $2\pi$ along the orbit, the oscillatory behavior integrates to
zero and \eqnname~(\ref{eq:jjtint}) simplifies to
\begin{equation}\label{eq:aAIJ}
  J_i = \frac{\int \dd \theta_i^A J_i^A}{\int \dd \theta_i^A}\,,
\end{equation}
that is, the target action is obtained by averaging the auxiliary
action over the auxiliary angle.

@@

Given the orbit integration already performed for calculating the
actions, we can use \equationname~(\ref{eq:aat}) to calculate the
frequencies and angles corresponding to the point $(\vecx,\vecv)$
(this procedure is similar to that used in \citealt{McMillan08a}). We
first re-write \equationname~(\ref{eq:aat}) in terms of the angle at
time zero and the frequencies
\begin{equation}\label{eq:aat2}
  \veca = \veca(t=0) + \veco(\vecj) \,t = \veca^A +2\sum_{\vecn > 0} \frac{\partial S_{\vecn}(\vecj)}{\partial \vecj}\,\sin(\vecn\cdot\veca^A)\,.
\end{equation}
This a linear system of equations for each point along the orbit with
unknowns $\veca(t=0)$, $\veco(\vecj)$, and the functions
$\frac{\partial S_{\vecn}(\vecj)}{\partial \vecj}$. For a finely
integrated orbit, this system is highly over-constrained (at least
when the number of expansion terms is small) and its solution can be
found by simple linear algebra for any desired number of expansion
terms $\sin(\vecn\cdot\veca^A)$. For extra stability, I integrate the
orbit both forward and backward in time, such that the desired angle
$\veca(t=0)$ is at the center of the dependent variable $t$.

@@

The method described here is completely general, but the only obstacle
to using it is that a sufficiently close isochrone potential has to be
supplied. Sufficiently close means specifically that the auxiliary
angles have to go through the full range of $0$ to $2\pi$ along the
orbit in the target potential. Whether they do or not is easy to check
and I have found that with limited trial and error a good enough
isochrone potential can be determined quickly for Milky-Way-like
axisymmetric potentials, although it remains to be checked whether
this holds for triaxial potentials as well. For analyzing stream data,
the necessity of this trial-and-error determination of the auxiliary
potential does not pose a practical problem, as all of the stream
stars are on very similar orbits. Therefore, the same auxiliary
potential can be used for all action--angle calculations for a given
potential and for different potentials as well, as long as they are
not too different. More widespread use of this technique may require
automated methods for determining a good auxiliary potential. In the
unlikely circumstance that no good auxiliary isochrone potential can
be found, the method discussed here can also be used with other
auxiliary potentials for which the actions and angles can be
calculated, such as other spherical potentials or the family of
St\"{a}ckel potentials. This will, however, add a non-negligible
amount of computation for the necessary numerical integrations.

>>>>

Recently, Fox \& Binney (2014, in preparation) have suggested a new
method for calculating $\vecj(\vecx,\vecv)$ inspired by torus
modeling. This method requires only a single orbit integration and
allows the actions to be calculated accurately and as precisely as
desired (that is, a convergence criterion can be applied and
convergence can be attained). I describe the Fox \& Binney method
briefly here and then show how to extend it to also calculate the
frequencies and angles: $(\vecx,\vecv) \rightarrow (\veco,\veca)$.

@@

The Fox \& Binney method consists of averaging \eqnname~(\ref{eq:jjt})
over the auxiliary angle along the path of the orbit, such that the
mapping defined by the $S_{\vecn}$ is onto orbital tori of the target
Hamiltonian, the oscillating cosine behavior cancels, and the target
action is obtained as the auxiliary-angle-averaged auxiliary
action. That is,
\begin{equation}\label{eq:jjtint}
  \int \dd \theta_i^A J_i^A = \int \dd \theta_i^A J_i + 2 \sum_{\vecn > 0} \vecn\,S_{\vecn}(J_i)\int \dd \theta_i^A\cos(\vecn\cdot\veca^A)\,,
\end{equation}
where the integral is over auxiliary angles calculated on the path of
the orbit in the target potential. The factor $\vecn\,S_{\vecn}(J_i)$
comes out of the integral as the target actions are conserved along
the orbit. In this Equation, we integrate over the angle coordinate
$\theta_i^A$ associated with the action $J_i$ that we want to compute
(that is, $\theta_R$ for $J_R$ and so on). This is because of the
factor $\vecn$: the condition $\vecn > 0$ means that when calculating
$J_R$ there are terms with $\vecn = (n_R,0,0)$, which would not cancel
according to the argument below when integrating over $\theta_Z$. For
non-resonant orbits and after a long-enough orbit integration, the
angles can be considered to be independent (that is, the orbit ``fills
its torus''), such that we can write (for example, for $\theta_R$)
\begin{equation}
\begin{split}
  \int \dd \theta_R^A\cos(\vecn\cdot\veca^A) & = \mathrm{Re}\left[\int \dd \theta_R^A\,\exp(-i\vecn\cdot\veca^A)\right]\,, \\
  & \approx \mathrm{Re}\left[\exp(-i\,n_\phi\,\theta^A_\phi-i\,n_Z\,\theta^A_Z)\int \dd \theta_R^A\,\exp(-i\,n_R\,\theta^A_R)\right]\,.
\end{split}
\end{equation}
As long as the auxiliary angle $\theta^A_R$ goes through the full range
$0$ to $2\pi$ along the orbit, the oscillatory behavior integrates to
zero and \eqnname~(\ref{eq:jjtint}) simplifies to
\begin{equation}\label{eq:aAIJ}
  J_i = \frac{\int \dd \theta_i^A J_i^A}{\int \dd \theta_i^A}\,,
\end{equation}
that is, the target action is obtained by averaging the auxiliary
action over the auxiliary angle.

@@

Given the orbit integration already performed for calculating the
actions, we can use \equationname~(\ref{eq:aat}) to calculate the
frequencies and angles corresponding to the point $(\vecx,\vecv)$
(this procedure is similar to that used in \citealt{McMillan08a}). We
first re-write \equationname~(\ref{eq:aat}) in terms of the angle at
time zero and the frequencies
\begin{equation}\label{eq:aat2}
  \veca = \veca(t=0) + \veco(\vecj) \,t = \veca^A +2\sum_{\vecn > 0} \frac{\partial S_{\vecn}(\vecj)}{\partial \vecj}\,\sin(\vecn\cdot\veca^A)\,.
\end{equation}
This is a linear system of equations for each point along the orbit
with unknowns $\veca(t=0)$, $\veco(\vecj)$, and the functions
$\frac{\partial S_{\vecn}(\vecj)}{\partial \vecj}$. For a finely
integrated orbit, this system is highly over-constrained (at least
when the number of expansion terms is small) and its solution can be
found by simple linear algebra for any desired number of expansion
terms $\sin(\vecn\cdot\veca^A)$. For extra stability, I integrate the
orbit both forward and backward in time, such that the desired angle
$\veca(t=0)$ is at the center of the dependent variable $t$.

@@

The method described here is completely general and the only obstacle
to using it is that a sufficiently close isochrone potential has to be
supplied. Sufficiently close means specifically that the auxiliary
angles have to go through the full range of $0$ to $2\pi$ along the
orbit in the target potential. Whether they do or not is easy to check
and I have found that with limited trial and error a good enough
isochrone potential can be determined quickly for Milky-Way-like
axisymmetric potentials in regions where the flattening is mild (\ie,
in the halo). Whether this method works near the plane of the disk
where the potential is strongly flattened remains to be checked,
although the fact that disk orbits are close to circular makes the
details of the auxiliary potential important. Similarly, the method
should be tested in more detail for triaxial potentials as well. For
analyzing stream data, the necessity of this trial-and-error
determination of the auxiliary potential does not pose a practical
problem, as all of the stream stars are on very similar
orbits. Therefore, the same auxiliary potential can be used for all
action--angle calculations for a given potential and for different
potentials as well, as long as they are not too different. More
widespread use of this technique may require automated methods for
determining a good auxiliary potential. In the unlikely circumstance
that no good auxiliary isochrone potential can be found, the method
discussed here can also be used with other auxiliary potentials for
which the actions and angles can be calculated, such as other
spherical potentials or the family of St\"{a}ckel potentials. This
will, however, add a non-negligible amount of computation for the
necessary numerical integrations.
